---
title: "Cyclistics-BusinessCase"
author: "Julio Venero"
date: "6/9/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Cyclistic Exercise Full Year Analysis

This analysis is for case study 1 from the Google Data Analytics Certificate (Cyclistic).

It’s originally based on the case study "'Sophisticated, Clear, and Polished’: Divvy and Data Visualization" written by Kevin Hartman [found here](https://artscience.blog/home/divvy-dataviz-case-study). 

We will be using the Divvy dataset for the case study. The purpose of this script is to consolidate downloaded Divvy data into a single dataframe and then conduct simple analysis to help answer the key question: *“In what ways do members and casual riders use Divvy bikes differently?”*

#### Packages used for the cleaning, analysis and visualization

* tidyverse for data import and wrangling
* lubridate for date functions
* ggplot for visualization
* janitor for data cleaning
* hms for time variables handling

```{r load libraries, include=FALSE}
library(tidyverse)  #helps wrangle data
library(lubridate)  #helps wrangle date attributes
library(ggplot2)  #helps visualize data
library(janitor)
library(hms)
```

### STEP 1: COLLECT DATA

Upload Divvy data sets (csv files) into the separate data frames
Last 12 months, from 05/2020 to 04/2021

```{r upload data sets, echo=TRUE, message=FALSE, warning=FALSE, results='hide'}
# Loop through csv files in directory and load to corresponding data frames
for (data in list.files(pattern="*.csv")) {
   name <- paste("tripdata", str_sub(data, 1, 6), sep = "_")
   assign(name, read_csv(data))
}
# Remove temporary variables created in the loop
rm(data)
rm(name)
```

### STEP 2: WRANGLE DATA AND COMBINE INTO A SINGLE FILE

Check consistency of columns among data frames

```{r variables check, echo=TRUE, message=FALSE}
# Janitor package has useful functions to compare columns in different data frames

# Returns mismatch columns
compare_df_cols(
    tripdata_202005,
    tripdata_202006,
    tripdata_202007,
    tripdata_202008,
    tripdata_202009,
    tripdata_202010,
    tripdata_202011,
    tripdata_202012,
    tripdata_202101,
    tripdata_202102,
    tripdata_202103,
    tripdata_202104,
    return = "mismatch"
)
```

The columns are the same but "end_station_id" and "start_station_id" have different data types: numeric and character.
We need all of them to be numeric

```{r convert data type, echo=TRUE, message=FALSE, warning=FALSE, results='hide'}
# Convert to numeric
tripdata_202012 <- tripdata_202012 %>%
  mutate(start_station_id = as.numeric(start_station_id),
         end_station_id = as.numeric(end_station_id))

tripdata_202101 <- tripdata_202101 %>%
  mutate(start_station_id = as.numeric(start_station_id),
         end_station_id = as.numeric(end_station_id))

tripdata_202102 <- tripdata_202102 %>%
  mutate(start_station_id = as.numeric(start_station_id),
         end_station_id = as.numeric(end_station_id))

tripdata_202103 <- tripdata_202103 %>%
  mutate(start_station_id = as.numeric(start_station_id),
         end_station_id = as.numeric(end_station_id))

tripdata_202104 <- tripdata_202104 %>%
  mutate(start_station_id = as.numeric(start_station_id),
         end_station_id = as.numeric(end_station_id))

```

Now, there should be no mismatch columns

```{r}
compare_df_cols(
    tripdata_202005,
    tripdata_202006,
    tripdata_202007,
    tripdata_202008,
    tripdata_202009,
    tripdata_202010,
    tripdata_202011,
    tripdata_202012,
    tripdata_202101,
    tripdata_202102,
    tripdata_202103,
    tripdata_202104,
    return = "mismatch"
)
```

Bind all data frames into a single one

```{r bind dataframes}
# Binding
tripdata <- bind_rows(
  tripdata_202005,
  tripdata_202006,
  tripdata_202007,
  tripdata_202008,
  tripdata_202009,
  tripdata_202010,
  tripdata_202011,
  tripdata_202012,
  tripdata_202101,
  tripdata_202102,
  tripdata_202103,
  tripdata_202104
)

# Remove the partial datasets
rm(tripdata_202005)
rm(tripdata_202006)
rm(tripdata_202007)
rm(tripdata_202008)
rm(tripdata_202009)
rm(tripdata_202010)
rm(tripdata_202011)
rm(tripdata_202012)
rm(tripdata_202101)
rm(tripdata_202102)
rm(tripdata_202103)
rm(tripdata_202104)

```

Create a column called "ride_length" equal to Started_at - Ended_at and a column called "day_of_week" to get the date that each ride started

```{r enrich data}
tripdata <- tripdata %>%
  mutate(ride_length = as_hms(difftime(ended_at, started_at)),
         day_of_week = wday(started_at, week_start = 1))
```


### STEP 3: CLEAN UP AND ADD DATA TO PREPARE FOR ANALYSIS

Inspect the new table that has been created

```{r data overview}
colnames(tripdata)  #List of column names
nrow(tripdata)  #How many rows are in data frame?
dim(tripdata)  #Dimensions of the data frame?
head(tripdata)  #See the first 6 rows of data frame.
str(tripdata)  #See list of columns and data types
summary(tripdata)  #Statistical summary of data.
```

#### 1. Inspect the values of each variable to identify duplicates, inconsistencies or errors

```{r variables inspect}
# Values consistency
unique(tripdata['rideable_type'])
unique(tripdata['start_station_name'])
unique(tripdata['end_station_name'])
unique(tripdata['member_casual'])

# Look for duplicated ride ids
duplicatedRideId <- tripdata[duplicated(tripdata$ride_id), "ride_id"]
nrow(duplicatedRideId) #209

duplicatedRides <- arrange(subset(tripdata, ride_id %in% duplicatedRideId$ride_id), ride_id)


```

Duplicated Ride IDs were found

Duplicated rides had a start date after the end date (negative ride length).

These rows will be deleted

```{r delete duplicates}
#Backup dataframe
tripdata_bk <- tripdata

#Remove duplicates
tripdata <- tripdata %>%
  filter(!(ride_id %in% duplicatedRideId$ride_id & ride_length < 0))

```

Check if there are more negative ride lengths and *remove them*

```{r remove negative ride length}
tripdata <- tripdata %>%
  filter(ride_length >= 0)
```

#### 2. Enrich the data frame

Define the correct data types for each variable

```{r redefine data types}
tripdata$rideable_type = as.factor(tripdata$rideable_type)
tripdata$member_casual = as.factor(tripdata$member_casual)
```

Add year, month, day columns to facilitate the analysis

```{r enrich with date columns}
tripdata$date <- as.Date(tripdata$started_at)
tripdata$day <- day(tripdata$date)
tripdata$month <- month(tripdata$date, label = TRUE)
tripdata$year <- year(tripdata$date)
```

#### 3. Remove rows based on business rules or with missing values (impute if possible)

The dataframe includes a few hundred entries when bikes were taken out of docks and checked for quality by Divvy. Remove them.

```{r remove quality check bikes}
tripdata <- tripdata[!tripdata$start_station_name == "HQ QR", ]
```

Check for missing data

```{r missing data}
sum(is.na(tripdata$ride_id)) #148183
tripdata <- tripdata[!is.na(tripdata$ride_id), ]
sum(is.na(tripdata$ride_id)) #0

sum(is.na(tripdata$start_station_id)) #439815
sum(is.na(tripdata$end_station_id)) #504254

```

Fill the missing station's id with values from other rows
```{r fill missing values}
# Create master table of start stations
station_name_mt <- tripdata %>%
  filter(!is.na(start_station_id)) %>%
  distinct(start_station_id, start_station_name) %>%
  arrange(start_station_id)
# Fill missing Start Station IDs
tripdata <- tripdata %>%
  left_join(station_name_mt, by="start_station_name", suffix = c("_td", "_mt")) %>%
  mutate(start_station_id_td = if_else(is.na(start_station_id_td), start_station_id_mt, start_station_id_td)) %>%
  select(-start_station_id_mt) %>%
  rename(start_station_id = start_station_id_td)

# Create master table of end stations
station_name_mt <- tripdata %>%
  filter(!is.na(end_station_id)) %>%
  distinct(end_station_id, end_station_name) %>%
  arrange(end_station_id)
# Fill missing End Station IDs
tripdata <- tripdata %>%
  left_join(station_name_mt, by="end_station_name", suffix = c("_td", "_mt")) %>%
  mutate(end_station_id_td = if_else(is.na(end_station_id_td), end_station_id_mt, end_station_id_td)) %>%
  select(-end_station_id_mt) %>%
  rename(end_station_id = end_station_id_td)

# Validate
colnames(tripdata)
sum(is.na(tripdata$start_station_id)) #117
sum(is.na(tripdata$end_station_id) & !is.na(tripdata$end_station_name)) #217

# Remove rows where start_station_id is still missing since the number of NAs is low
tripdata <- subset(tripdata, !is.na(start_station_id))
```

### STEP 4: CONDUCT DESCRIPTIVE ANALYSIS

#### 1. Descriptive statistics

```{r descriptive analysis}
summary(tripdata)

aggregate(tripdata$ride_length ~ tripdata$member_casual, FUN = mean)
aggregate(tripdata$ride_length ~ tripdata$member_casual, FUN = median)
aggregate(tripdata$ride_length ~ tripdata$member_casual, FUN = max)
aggregate(tripdata$ride_length ~ tripdata$member_casual, FUN = min)

```

Compare members and casual users ride length statistics

```{r members vs casual, warning=FALSE}
# Members and casual users statistics
tripdata %>%
  group_by(member_casual) %>%
  summarise(ride_length_avg = as_hms(mean(ride_length)),
            ride_length_med = as_hms(median(ride_length)),
            ride_length_max = as_hms(max(ride_length)),
            ride_length_min = as_hms(min(ride_length))
          )

```

Add a new column to have the name of the weekdays and analyze ride_length by it

```{r member vs casual by weekday, warning=FALSE}
# Add weekday column
tripdata <- tripdata %>%
  mutate(weekday = as.factor(weekdays(started_at)))

# Order weekday levels: from monday to sunday
tripdata$weekday <- factor(tripdata$weekday,
                           levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday",
                                      "Saturday", "Sunday"))

# Members and casual users statistics by weekday
tripdata %>%
  group_by(member_casual, weekday) %>%
  summarise(
            number_of_rides = n(),
            average_duration = as_hms(mean(ride_length)),
            median_duration = as_hms(median(ride_length)),
            max_duration = as_hms(max(ride_length)),
            min_duration = as_hms(min(ride_length))
          )

```

#### 2. Analyze through visualizations 

*Analysis by weekday*

Let's visualize the number of rides by rider type

```{r rides per weekday viz, message=FALSE, warning=FALSE}
tripdata %>% 
  group_by(member_casual, weekday) %>%
  summarise(number_of_rides = n()) %>%
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge") +
  theme(axis.title.x = element_blank()) +
  ylab("Number of rides") +
  scale_y_continuous(labels = scales::comma) +
  labs(fill = "Customer Type",
       title = "Rides per weekday",
       subtitle = "Members' number of rides is higher, except for the weekends") +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.line = element_line(colour = "black"))
```

Now, check the average duration

```{r avg duration per weekday viz, message=FALSE, warning=FALSE}
tripdata %>%
  group_by(member_casual, weekday) %>%
  summarise(average_duration = as_hms(mean(ride_length))) %>%
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge") +
  theme(axis.title.x = element_blank()) +
  ylab("Average Duration") +
  labs(fill = "Customer Type",
       title = "Customer's Average ride duration per weekday",
       subtitle = "Duration remains similar throughout the week.\nCasual riders duration is double that of members") +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.line = element_line(colour = "black"))

```

Start time per weekday

```{r, message=FALSE, warning=FALSE}
# Create the function to calculate the mode.
# Source: https://www.tutorialspoint.com/r/r_mean_median_mode.htm
getmode <- function(v) {
  uniqv <- unique(v)
  uniqv[which.max(tabulate(match(v, uniqv)))]
}

tripdata %>%
  group_by(member_casual, weekday) %>%
  summarise(started_at_hour = hour(getmode(started_at))) %>%
  ggplot(aes(x = weekday, y = started_at_hour, color = member_casual, group = member_casual)) +
  geom_line() +
  geom_point() +
  theme(axis.title.x = element_blank()) +
  ylab("Ride Start Hour") +
  labs(fill = "Customer Type",
       title = "Most frequent hour to start a ride",
       subtitle = "Similar behavior except for monday") +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.line = element_line(colour = "black"))

```

*Analysis by month*

Let's visualize the number of rides by rider type

```{r rides per month viz, message=FALSE, warning=FALSE}
tripdata %>% 
  group_by(member_casual, month) %>%
  summarise(number_of_rides = n()) %>%
  ggplot(aes(x = month, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge") +
  theme(axis.title.x = element_blank()) +
  ylab("Number of rides") +
  scale_y_continuous(labels = scales::comma) +
  labs(fill = "Customer Type",
       title = "Rides per month",
       subtitle = "Similar number of rides around mid-year\nThe difference is higher at the beginning and end of the year") +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.line = element_line(colour = "black"))
```

Now, check the average duration

```{r avg duration per month viz, message=FALSE, warning=FALSE}
tripdata %>%
  group_by(member_casual, month) %>%
  summarise(average_duration = as_hms(mean(ride_length))) %>%
  ggplot(aes(x = month, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge") +
  theme(axis.title.x = element_blank()) +
  ylab("Average Duration") +
  labs(fill = "Customer Type",
       title = "Customer's Average ride duration per month",
       subtitle = "Casual riders duration is double that of members") +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.line = element_line(colour = "black"))

```

*Is there any difference regarding the type rideable type?*

```{r rideable type diff, message=FALSE, warning=FALSE}
tripdata %>%
  group_by(member_casual, rideable_type) %>%
  summarise(number_of_rides = n()) %>%
  ggplot(aes(x=rideable_type, y=number_of_rides)) +
  geom_col(aes(fill=member_casual), position = "dodge") +
  coord_flip() +
  theme(axis.title.y = element_blank()) +
  ylab("Rides") +
  scale_y_continuous(labels = scales::comma) +
  labs(fill = "Customer Type",
       title = "Customer rides by rideable type",
       subtitle = "No signicant differences between customers") +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.line = element_line(colour = "black"))

```


### STEP 5: EXPORT SUMMARY FILE FOR FURTHER ANALYSIS

Create a csv file that we will visualize in Excel, Tableau, or my presentation software

```{r export to csv}
counts <- aggregate(tripdata$ride_length ~ tripdata$member_casual + tripdata$weekday, FUN = mean)
write.csv(counts, 'avg_ride_length.csv')
```
